---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Assignment Instructions
1.Code for reading in the dataset and/or processing the data
2.Histogram of the total number of steps taken each day
3.Mean and median number of steps taken each day
4.Time series plot of the average number of steps taken
5.The 5-minute interval that, on average, contains the maximum number of steps
6.Code to describe and show a strategy for imputing missing data
7.Histogram of the total number of steps taken each day after missing values are imputed
8.Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
9.All of the R code needed to reproduce the results (numbers, plots, etc.) in the report


## Loading and preprocessing the data
```{r, echo = TRUE}
setwd(getwd())
activity<-read.csv("activity.csv")
```

Exploring the basics of this data
```{r}
dim(activity)
names(activity)
head(activity)
str(activity)
#total number of missing data
sum(is.na(activity$steps))/dim(activity)[[1]]
#transforming the date column into date format using lubridate
library(lubridate)
activity$date<-ymd(activity$date)
length(unique(activity$date))
```

## Step 2
## Histogram of the total number of steps taken each day
```{r, echo = TRUE}
library(ggplot2)
D2<-data.frame(tapply(activity$steps,activity$date,sum,na.rm=TRUE))
D2$date<-rownames(Q2)
rownames(Q2)<-NULL
names(D2)[[1]]<-"Total Steps"
png("plot1.png")
#Total Steps by date bar chart
ggplot(D2,aes(y=D2$`Total Steps`,x=D2$date))+geom_bar(stat="identity") + ylab("Total Steps")+xlab("Date")+ggtitle("Total Steps by date")
dev.off()
ggplot(D2,aes(y=D2$`Total Steps`,x=D2$date))+geom_bar(stat="identity") + ylab("Total Steps")+xlab("Date")+ggtitle("Total Steps by date")
#Histogram of total steps
qplot(D2$`Total Steps`,geom="histogram",xlab="Total Steps",ylab="Counts",main="Total Steps Historgram")
png("plot1.1.png")
qplot(D2$`Total Steps`,geom="histogram",xlab="Total Steps",ylab="Counts",main="Total Steps Historgram")
dev.off()
```

## Step 3
## Mean and median number of steps taken each day
```{r, echo = TRUE}
library(dplyr)
D3<-data.frame(round(tapply(activity$steps,activity$date,mean,na.rm=TRUE),2))
D3$date<-rownames(D3)
rownames(D3)<-NULL
names(D3)[[1]]<-"Mean Steps"
temp<-activity%>%select(date,steps) %>% group_by(date) %>% summarise(median(steps))
names(temp)[[2]]<-"Median Steps"
D3$median<-temp$`Median Steps`
D3<-D3 %>% select(date,`Mean Steps`,median)
```


## Step 4
## Time series plot of the average number of steps taken
```{r, echo = TRUE}
D4<-D3
D4$date<-as.Date(D4$date,format="%Y-%m-%d")
ggplot(D4,aes(x=D4$date,y=D4$`Mean Steps`))+geom_bar(stat="identity")+scale_x_date()+ylab("Mean Steps Every day")+xlab("Date")+ggtitle("Mean Steps by Date")
png("plot4.png")
ggplot(D4,aes(x=D4$date,y=D4$`Mean Steps`))+geom_bar(stat="identity")+scale_x_date()+ylab("Mean Steps Every day")+xlab("Date")+ggtitle("Mean Steps by Date")
dev.off()
```

## Step 5
## The 5-minute interval that, on average, contains the maximum number of steps

```{r, echo = TRUE}
#This is assuming that the words on average means averaging steps by date and interval
activity$interval<-factor(activity$interval)
D5<-aggregate(data=activity,steps~date+interval,FUN="mean")
D5<-aggregate(data=D5,steps~interval,FUN="max")
```

## Step 6
Code to describe and show a strategy for imputing missing data
There are multiple strategies to deal with multiple value imputations.
The common strategies include:
1. Constant value imputations
2. Regression model value imputations
3. Mean/mode value substitutions
For the purpose of simplicity, in this question, I will use the mean/mode value substitution strategy to impute missing values. That is, using the mean values to substitute out the missing values in the original data set
Before doing any sort of imputation, it is helpful to understand what are the distributions of missing values by date and interval
```{r, echo = TRUE}
D6<-activity
D6$Missing<-is.na(D6$steps)
D6<-aggregate(data=D6,Missing~date+interval,FUN="sum")
D6.1<-data.frame(tapply(D6$Missing,D6$date,sum))
D6.1$date<-rownames(D6.1)
rownames(D6.1)<-NULL
names(D6.1)<-c("Missing","date")
D6.1$date<-as.Date(D6.1$date,format="%Y-%m-%d")
D6.2<-data.frame(tapply(D6$Missing,D6$interval,sum))
D6.2$date<-rownames(D6.2)
rownames(D6.2)<-NULL
names(D6.2)<-c("Missing","Interval")
par(mfrow=c(1,2))
plot(y=D6.1$Missing,x=D6.1$date,main="Missing Value Distribution by Date")
plot(y=D6.2$Missing,x=D6.2$Interval,main="Missing Value Distribution by Interval")
table(activity$date)
```
